---
title: "factorMerger: a set of tools to support results from post-hoc testing"
author: "Agnieszka Sitko"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{factorMerger: a set of tools to support results from post hoc testing}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE, message = FALSE,
                      fig.height = 5, fig.width = 10)
```

# Introduction

The aim of `factorMerger` is to provide a set of tools to support results from post hoc comparisons. Post hoc testing is an analysis performed after running *ANOVA* to examine differences between group means (of some response numeric variable) for each pair of groups (groups are defined by a factor variable).

This project arose from the need to create a method of post hoc testing which gives the hierarchical interpretation of relations between groups means. Thereby, for a given significance level we may divide groups into non-overlapping clusters. 

# Algorithm inputs

In the current version the **factorMerger** package supports parametric models: 

- one-dimensional Gaussian (with the argument `family = "gaussian"`),
- multi dimensional Gaussian (with the argument `family = "gaussian"`),
- binomial (with the argument `family = "binomial"`),
- survival (with the argument `family = "survival"`).

There are four algorithms available: adaptive, fast-adaptive, fixed and fast-fixed (they are set with the `method` argument of `mergeFactors`). *Fast* algorithms enable to unite only those groups whose group statistics (i.e. means in the Gaussian case) are close.

# Generating samples 

To visualize functionalities of `factorMerger` we use samples or real data examples with response variable whose distribution follow one of the listed above. The corresponding factor variable is sampled uniformly from a finite set of a size $k$. 

To do so, we may use function `generateSample` or `generateMultivariateSample`.

```{r}
library(factorMerger) 
library(knitr)
library(dplyr)
randSample <- generateMultivariateSample(N = 100, k = 10, d = 3)
```

# Merging factors

`mergeFactors` is a function that performs hierarchical post hoc testing. As arguments it takes:

 - matrix/data.frame/vector with numeric response,
 - factor vector defining groups.
 
By default (with argument `abbreviate = TRUE`) factor levels are abbreviated and surrounded with brackets.

## Multi-dimensional Gaussian model


### Computations

```{r}
fmAll <- mergeFactors(randSample$response, randSample$factor)
```

`mergeFactors` outputs with information about the 'merging history'. 

```{r}
mergingHistory(fmAll, showStats = TRUE) %>% 
    kable()
```

Each row of the above frame describes one step of the merging algorithm. First two columns specify which groups were merged in the iteration, columns *model* and *GIC* gather loglikelihood and Generalized Information Criterion for the model after merging. Last two columns are p-values for the *Likelihood Ratio Test* -- against the full model (*pvalVsFull*) and against the previous one (*pvalVsPrevious*).

If we choose a *fast* version of merging one dimensional response is fitted using `isoMDS{MASS}`. Next, in each step only groups whose means are closed are compared.

```{r}
fm <- mergeFactors(randSample$response, randSample$factor, 
                   method = "fast-fixed")

mergingHistory(fm, showStats = TRUE) %>% 
    kable()
```

### Final clusters

Algorithms implemented in the **factorMerger** package enable to create unequivocal partition of a factor. Below we present how to extract the partition from the `mergeFactor` output.

 - predict new labels for observations
 
```{r}
cutTree(fm)
```
By default, `cutTree` returns a factor split for the optimal GIC (with penalty = 2) model. However, we can specify different metrics (`stat = c("loglikelihood", "p-value", "GIC"`) we would like to use in cutting. If `loglikelihood` or `p-value` is chosen an exact threshold must be given as a `value` parameter. Then `cutTree` returns factor for the smallest model whose statistic is higher than the threshold. If we choose `GIC` then `value` is interpreted as GIC penalty.

```{r}
mH <- mergingHistory(fm, T)
thres <- mH$model[nrow(mH) / 2]
cutTree(fm, stat = "loglikelihood", value = thres)
```
In this example data partition is created for the last model from the merging path whose loglikelihood is greater than `r thres`.

 - get final clusters and clusters dictionary
 
```{r}
getOptimalPartition(fm)
```

Function `getOptimalPartition` returns a vector with the final cluster names from the factorMerger object. 

```{r}
getOptimalPartitionDf(fm)
```

Function `getOptimalPartitionDf` returns a dictionary in a data frame format. Each row gives an original label of a factor level and its new (cluster) label.


Similarly to `cutTree`, functions `getOptimalPartition` and `getOptimalPartitionDf` take arguments `stat` and `threshold`.

### Visualizations

We may plot results using function `plot`. 


```{r, fig.height = 9, fig.width = 18}
plot(fm, panel = "all", nodesSpacing = "equidistant", colorCluster = TRUE)
```

```{r, fig.height = 5, fig.width = 10}
plot(fmAll, panel = "tree", statistic = "p-value", 
     nodesSpacing = "effects", colorCluster = TRUE)
```

```{r}
plot(fm, colorCluster = TRUE, panel = "response")
```

The heatmap on the right shows means of all variables taken into analysis by groups.



```{r, fig.height = 5, fig.width = 10}
plot(fm, colorCluster = TRUE, panel = "response", responsePanel = "profile")
```

In the above plots colours are connected with the group. The plot on the right shows means rankings for all variables included in the algorithm. 


It is also possible to plot *GIC* together with the merging path plot.

```{r, fig.width = 5}
plot(fm, panel = "GIC", penalty = 5)
```

Model with the lowest GIC is marked.

## One-dimensional Gaussian model

```{r}
oneDimRandSample <- generateSample(1000, 10)
```
```{r}
oneDimFm <- mergeFactors(oneDimRandSample$response, oneDimRandSample$factor, 
                         method = "fixed")
mergingHistory(oneDimFm, showStats = TRUE) %>% 
    kable()
```
```{r}
plot(oneDimFm, palette = "Reds")
```

```{r}
plot(oneDimFm, responsePanel = "boxplot", colorCluster = TRUE)
```

## Binomial model


If `family = "binomial"` response must have to values: `0` and `1` (`1` is interpreted as success).

```{r}
binomRandSample <- generateSample(1000, 10, distr = "binomial")
table(binomRandSample$response, binomRandSample$factor) %>% 
    kable()
```

```{r}
binomFm <- mergeFactors(binomRandSample$response, 
                        binomRandSample$factor, 
                        family = "binomial", 
                        method = "fast-adaptive")
mergingHistory(binomFm, showStats = TRUE) %>% 
    kable()
```

```{r}

```

```{r}
plot(binomFm, colorCluster = TRUE, penalty = 7)
```

```{r}
plot(binomFm, gicPanelColor = "red")
```


## Survival model
 
If `family = "survival"` response must be of a class `Surv`.
 
```{r}
 library(survival)
 data(veteran)
 survResponse <- Surv(time = veteran$time, 
                  event = veteran$status)
 survivalFm <- mergeFactors(response = survResponse, 
                    factor = veteran$celltype, 
                    family = "survival")
```

```{r}
 mergingHistory(survivalFm, showStats = TRUE) %>%
     kable()
```


```{r}
 plot(survivalFm)
```

```{r}
 plot(survivalFm, nodesSpacing = "effects", colorCluster = TRUE)
```

